Conceptual and methodological problems with bias detection and avoidance in natural language processing

### 2021-06-08 thesis/_book/_main.tex
added chapter 2

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-2-1} \end{center}

@@

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-6-1} \end{center}

@@

@@

@@

@@

@@

@@

@@

@@

@@

\hypertarget{refs}{}
\hypertarget{ref-Caliskan2017Semantics}{}
Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
automatically from language corpora necessarily contain human biases.
\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
\url{http://arxiv.org/abs/1608.07187}

>>>>

\chapter{Cosine similarity and bias
detection}\label{cosine-similarity-and-bias-detection}

\section{Word embeddings}\label{word-embeddings}

To understand what cosine similarity measurement is, one first needs to
grasp the concept of translating words to a computer-readable form. In
the field of natural language processsing there are two main types of
words representation --- localist and distributed. One-hot encoding is
an example of a method used to achieve a localist representation of
words. Here each vector contains information only about a single data
point, this is achieved by first mapping categorical values (words) to
integers and then to each integers a binary vector is assigned which
contains only 0s except for the index of the integer, which is assigned
1. An example of a localist representation is:

\begin{longtable}[]{@{}llllll@{}}
\toprule
word & 1 & 2 & 3 & 4 & 5\tabularnewline
\midrule
\endhead
woman & 1 & 0 & 0 & 0 & 0\tabularnewline
man & 0 & 1 & 0 & 0 & 0\tabularnewline
girl & 0 & 0 & 1 & 0 & 0\tabularnewline
boy & 0 & 0 & 0 & 1 & 0\tabularnewline
monarch & 0 & 0 & 0 & 0 & 1\tabularnewline
\bottomrule
\end{longtable}

In the example above it is clear that the length of the vectors
increases with the number of words in a vocabulary. It is not a very
computationally efficient representation. It has other flaws as well.
For example, it is unable to capture the resemblance between words
appearing in similar contexts.

In contrast to the localist representation, a distributed representation
returns vectors that contain continuous values instead of discrete 1s
and 0s. Word embeddings are a class of various techniques that allow one
to represent words as distributed vectors. Such learned representations
of text have certain properties. At least prima facie, they store
similar (or at least co-occurring) words close to each other in a vector
space. An example of distributed representation is:

\begin{longtable}[]{@{}lllll@{}}
\toprule
word & 1 & 2 & 3 & 4\tabularnewline
\midrule
\endhead
woman & 0.456 & 0.267 & 0.675 & 0.131\tabularnewline
man & 0.451 & 0.897 & 0.472 & 0.088\tabularnewline
girl & 0.604 & 0.262 & 0.414 & 0.706\tabularnewline
boy & 0.279 & 0.172 & 0.475 & 0.010\tabularnewline
monarch & 0.565 & 0.678 & 0.463 & 0.975\tabularnewline
\bottomrule
\end{longtable}

One of the advantages of using a distributed representation is that one
is able to represent an enormous number of concepts with a smaller
number of units. It is also possible to better capture similarities as
words of similar meanings can have similar numeric vectors.

Word embeddings have many applications in natural language processing.
They are handy in document search and information retrieval. They also
play their part in improving automatic translations. Well learned word
representations may also contribute to the improvement of sentiment
analysis or spam detection.

The numbers occurring in such representations are not random. They are
learned in a process that uses a very shallow neural network. There are
various types of techniques used for learning the vectors
representations. One of the most straightforward ones is a skip-gram
model. Given a word the models tries to predict its neighboring words
from the sentence. The mathematics behind the process relies on the idea
that the prediction concerns the conditional probability of the adjacent
words. The algorithm tries to minimize the loss function, which
penalizes the system for discrepancy with actual co-occurrence
frequencies in the corpus. One can choose various parameters of the
model, such as the window size that determines how many surrounding
words the model should predict. After preparing such a fitted model one
takes only the learned weights from a neural network, and uses them as
vectors in a word embeddings representation.

\section{Cosine similarity and
distance}\label{cosine-similarity-and-distance}

Cosine similarity is often used as a method of finding out whether
vector representations for two words suggest that they are similar or
somehow connected. Cosine similarity is the cosine of the angle between
two vectors: the result of dividing their inner product (dot product
usually) by the product of their magnitudes.

\begin{align} \tag{Sim}
\mathsf{cosineSimilarity}(A,B) & = \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert}
\end{align}

Cosine similarity is considered a proper tool for this operation as its
result has a clear connection to geometry and at least for a low number
of dimensions may be easily interpreted. Using this scale, one can
compare vector similarities in a fairly clear manner. When the vectors
are aligned perpendicularly to each other, their similarity equals 0
(which is the same as the cosine of 90 degrees). This tells us that the
similarity between the vectors is small. As the angle between vectors
decreases, cosine similarity approaches one, which stands for the
greatest similarity.

One of the limitations of this measure is that it informs us only about
similarities between vectors in terms of their orientation. However, it
is often argued that in comparing words in terms of this metric, the
magnitude of vectors may be treated as irrelevant, as the most important
information pertains to direction. In what follows, it is important to
distinguish between cosine similarity and cosine distance, defined as:

\begin{align} \tag{Sim}
\mathsf{cosineDistance}(A,B) &  = 1 - \mathsf{cosineSimilarity}(A,B)\\
 &  = 1 - \frac{A \cdot B}{\vert \vert A \vert \vert \,\vert \vert B \vert \vert} \nonumber
\end{align}

The greater the similarity between two vectors, the smaller the distance
between them. The cosine distance ranges between 0 and 2. If the vectors
are in an opposite direction to each other, the cosine distance is 2.
And if the vectors are extremely similar then the cosine distance is
very close to 0.

It is worth mentioning one more point concerning cosine similarity.
After the vectors are normalized to have length equal to 1, inner
product itself (often dot product) is used to measure the similarity.

\section{Cosine distance in a one-class bias
detection}\label{cosine-distance-in-a-one-class-bias-detection}

Bolukbasi, Chang, Zou, Saligrama, \& Kalai
(\protect\hyperlink{ref-Bolukbasi2016Man}{2016}) define similarity
between words as the inner product of their normalized vectors (that is,
they use cosine similarity). They focus on examining what the geometry
of word embedding is in regard to ``he'' and ``she'' words. In other
words, whether the similarity between those concepts and other words
reflects expected gender stereotypes. They test this hypothesis by
investigating whether there is a connection between word embeddings
representing certain professions and words referring to gender. They
also evaluate whether automatically produced analogies between words
reflect the stereotypes as well.

A very vivid way to follow their method of arguing that bias in word
embeddings is real is to plot the values of inner product of chosen
words. The plot below does not originate from the original paper (it
comes from
\url{https://www.kaggle.com/rtatman/gender-bias-in-word-embeddings}) but
a similar visualization may be found there. Data used to create our plot
is as follows. \newline

Occupations associated with feminine:
\textbf{"homemaker", "nurse", "receptionist", "librarian", "socialite", "hairdresser", "nanny", "bookkeeper", "stylist", "housekeeper", "interior designer", "guidance counselor"}
\newline

Occupations associated with masculine:
\textbf{"maestro", "skipper", "protege", "philosopher", "captain", "architect", "financier", "warrior", "broadcaster", "magician", "fighter pilot", "boss"}
\newline

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-1-1} \end{center}

The points in the plot above result from the calculation of the inner
product of a chosen vector for a profession word and a vector for a
gender word (she or he). Inner product of two vectors expresses
similarity between words. This assumption originates from the geometry
and properties of a vector space.

\section{Cosine distance in a multi-class bias
detection}\label{cosine-distance-in-a-multi-class-bias-detection}

In the article Manzini, Lim, Tsvetkov, \& Black
(\protect\hyperlink{ref-manzini2019black}{2019}) the authors present
different approach towards finding similarities between classes of
words. The authors claim that texts available online are full of direct
or indirect human stereotypes. As a result, word embeddings are prone to
learn and maybe amplify those biases and propagate them further into AI
models that are used for various applications. Cosine distance is used
in the article as a measure to first argue for the existence of
multi-class bias and then to show how through bias mitigation techniques
the bias may be decreased. The main steps in the procedure are as
follows.

As the code has been provided by the authors we were able to reconstruct
their results. Let us go through an example that refers to the process
of hard debiasing on religious attributes.

The example refers to the process of hard debiasing on religious
attributes.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  First we load word embeddings from
  reddit.US.txt.tok.clean.cleanedforw2v.w2v dataset
\end{enumerate}

\begin{itemize}
\tightlist
\item
  The word embeddings have only 50 dimensions
\item
  The number of individual words from the dataset is 44895
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  The protected group the authors assume should ideally not have high
  cosine similarity to stereotypical words. The word embeddings geometry
  should not place this group close to harmful stereotypes, if it is to
  be bias-free.
\end{enumerate}

For instance, let's look at the religion-related words.

Protected words by religion type:

\textbf{"jew"       : ["judaism", "jew", "synagogue", "torah", "rabbi"]}
\newline
\textbf{"christian" : ["christianity", "christian", "church", "bible", "priest"]}
\newline
\textbf{"muslim"    : ["islam", "muslim", "mosque", "quran", "imam"]}
\newline

Stereotypical words by religion type:

\textbf{"jew"       : ["greedy", "cheap", "hairy", "liberal"]} \newline
\textbf{"christian"       : ["judgemental", "conservative", "familial"]}
\newline
\textbf{"muslim"       : ["violent", "terrorist", "dirty", "uneducated"]}
\newline

We have prepared a dataset including the values of cosine distance for
each protected word with each attribute (stereotype), its head is
displayed in the table.

\caption{\label{tab:religionTableHeadEarly}Head of the religion dataset.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lllrrl}
\toprule
protectedWord & wordToCompare & wordClass & cosineDistance & cosineSimilarity & connection\\
\midrule
judaism & violent & muslim & 0.7141939 & 0.2858061 & different\\
judaism & terrorist & muslim & 0.7461333 & 0.2538667 & different\\
judaism & dirty & muslim & 1.2002599 & -0.2002599 & different\\
judaism & uneducated & muslim & 0.7885469 & 0.2114531 & different\\
judaism & greedy & jewish & 1.0026172 & -0.0026172 & associated\\
judaism & cheap & jewish & 1.2323229 & -0.2323229 & associated\\
\bottomrule
\end{tabular}}
\end{table}

\pagebreak

In the article there was no analysis of individual distances but the
general look at the data through the usage of mean. The authors
introduced a metric that tries to generalize the presence of bias
through the classification of multi-class bias in groups of words
connected with gender, religion or race.

In the process they first take the mean of cosine distances between a
given protected word and each stereotype. They do not differentiate
between stereotypes associated with a word and stereotypes associated
with different words (in case of religion stereotypes characteristic for
christianity has also cosine distance measured with judaism or islam).
Then after collecting the list of mean cosine distances, they average
the list to get one final value represeting whole group, in this example
religion. In case of religion, the final mean of all mean distances is
equal to 0.859.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  In the article the authors also try to remove previously defined
  biases from word embedding.
\end{enumerate}

First they identify the bias subspace using Principal Component Analysis
(PCA) which is a technique for dimensionality reduction. It is applied
here to choose the subspace that contains the greatest amount of
information. There can be many subspaces found in a given group, for
example in terms of religion one can identify at least a few sets that
are to grasp the concept of religion in general: \newline
\textbf{["judaism", "christianity", "islam"]} \newline
\textbf{["jew", "christian", "muslim"]} \newline
\textbf{["synagogue", "church", "mosque"]} \newline

The idea is to find a set that provides enough information to create
from it a vector representing the concept of religion among words. This
strategy is based on the idea that different dimensions of vectors
contain different types of information and in some words in vector
layers (subspaces) the information about religiousness is implicitly
conveyed. In some cases this knowledge is useful but in case of harmful
stereotypes one does not want to have religion concept in stereotypical
words.

After finding bias subspace they use it to modify the vectors values
individually so that their cosine distances towards certain words are
changed. In case of stereotypes the aim is to make the cosine distances
larger so that the association between protected word and harmful
stereotype is smaller.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Finally there is a multi-class evaluation. They check each protected
  attribute with each group of debiased words. Then they do each time
  mean from the cosine distances from protected word to neutral words.
  Then they do a mean from all of these means to return one value. If
  the returned value of cosine distance is greater than the one
  calculated at the very beginning then it means that the bias was
  removed as the similarity between stereotypical word and the protected
  one decreased.
\end{enumerate}

How are the multi-class biases actually defined? The authors take a list
of alleged biases towards protected groups like gender, race, religion
from various scientific reports. The multi-class bias existence is
further proved by providing the reader with the value of a mean from the
cosine similarity between protected word and list of of previously
defined biases.

\chapter{Limitations of the approach}\label{limitations-of-the-approach}

\section{Selection and number of
attributes}\label{selection-and-number-of-attributes}

The attributes are taken from different sources, therefore there is no
one justification for their choice.

In the article there is no mention of methodology for deciding on the
number of attributes necessary to prove a hypothesis on the given size
of dataset. There are however some ways to estimate how many samples we
need to make sure that the result is significant.

\section{No control groups}\label{no-control-groups}

\textbf{Control group}
\todo{Tutaj raczej powiedz ze nie maj agrup a ze my bedziemy mieli}

How to properly prepare control group in terms of quality and quantity?

\section{Means of means}\label{means-of-means}

The authors use the mean average cosine similarity to check on
multi-class similarity between protected word and harmful stereotypes.
Details of the process are described as follows.

When one calculates individual cosine similarity between some chosen
word like ``christianity'' and neutral attributes that are supposed to
be biased are actually often having very small and negative similarity
value. As a result the measurement cosine distance is for such pair of
words greater than 1. The similarity interpretation for that would
suggest that the words are not similar as their cosine distance is so
enormous.

Doing a mean hides this issue and as there are pairs having negative and
small similarities and there are those that have similarity equal to
0.5, the resulting calculation seems to be in norm. Additionally in such
method the uncertainty interval is also not included which makes it even
more difficult to give reasonable interpretations of the results. In the
original paper words from all three religions were checked with all of
the stereotypes which means that there was no distinction on whether the
stereotype is associated with given religion or with the other. Not all
of the stereotypical words should be considered as harmful for all of
the religions. In our research we distinguished between associated
stereotypes, not associated (stereotype is valid but for different
religion than currently checked) and none (meaning neutral words).
\newline

\section{Interpretability issues}\label{interpretability-issues}

Assuming for a moment that the value of multi-class cosine distance is
correct, one may question the results' interpretation. In {[}Black is to
Criminal as Caucasian is to Police:Detecting and Removing Multiclass
Bias in Word Embeddings{]} Table 2, there are summarized the averages of
cosine distance per group (gender, race, religion). I would like to
focus now on analyzing the values relating to religious biases. Here is
the fragment of table that refers to that:

\begin{longtable}[]{@{}ll@{}}
\toprule
Religion Debiasing & MAC\tabularnewline
\midrule
\endhead
Biased & 0.859\tabularnewline
Hard Debiased & 0.934\tabularnewline
Soft Debiased (\(\lambda\) = 0.2) & 0.894\tabularnewline
\bottomrule
\end{longtable}

MAC stand for mean average cosine similarity although in reality the
values of cosine distance are there stored. What may attract attention
is the fact that the value of cosine distance in ``Biased'' category is
already quite high even before ``debiasing''. High cosine distance
indicates low cosine similarity between values. One could think that
average cosine similarity equal to approximately 0.141 is not
significant enough to consider it as biased. However the authors aim to
mitigate ``biases'' in vectors with such great distance to make it even
larger. Methodologically there is a question on what bases this small
similarity is still considered as a proof of bias presence.

This is in general the problem of scale and the lack of universal
intervals that could be applied to know whether the cosine similarity is
high enough to considered given words more similar that if we chose them
on random. In other metrics like Pearson coefficient we are provided
with artificial interval that tries to make the interpretations more
objective with a point of reference. It seems like similar practic could
be beneficial for metrics aimed at measuring similarity between words.

\section{The curse of dimensionality}\label{the-curse-of-dimensionality}

Curse of dimensionality may take place when there is an increase in
volume of data that results in adding extra dimensions to the Euclidean
space. According to the article
``\url{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}''
as the number of features increases, it may be harder and harder to get
useful information from the data with the usage of available algorithms.
One may notice that more data should contribute to greater amount of
information but more information also means greater risk of noise and
distractions in data. At the same time, many times modern solutions are
adapted to smaller dimensions and their results in higher ones are not
intuitive or may be prone to be mistaken.

Using cosine similarity in high dimensions in word embeddings may also
be prone to the curse of dimensionality. According to this article
``\url{https://www.researchgate.net/publication/327498046_The_Curse_of_Dimensionality_Inside_Out}''
there are reasons to consider this phenomenon when searching for word
similarities in higher dimensions.

In the article an experiment is conducted that aims at showing how the
similarity values and variation change as the number of dimensions
increases. The hypothesis made in the paper states that two things will
happen as the number of dimensions increase, the first one is that
effort required to measure cosine similarity will be greater and the
second one is that the similarity between data will blur out and have
less variation. In details, the authors generate random points with
increasing number of dimensions where each dimension of a data point is
given a value between 0 and 1. Then they pick one vector on random from
each dimension class and calculate cosine similarity between chosen
vector and the rest of the data. Then they check how the variation of
values changes as the number of dimensions increases. It seems like the
more dimensions there are, the smaller the variance and therefore it is
less obvious how to interpret the resulting cosine similarities. Maybe
the scale should be adjusted to the number of dimensions and variance so
that it still gives us sensible information about data. According to
some publishers the cosine similarity in high dimensions is not reliable
enough to trust it as it may be the case that choosing words on random
may result in getting similar values as when picking them consciously.
\newline

\begin{figure}
\centering
\includegraphics{../images/curseOfDimensionality.png}
\caption{curse of dimensionality, number of dimensions on the x axis,
standard deviation of similarity on the y axis}
\end{figure}

--\textgreater{}

--\textgreater{}

--\textgreater{}

--\textgreater{}

--\textgreater{}

\section{Loading and understanding the
dataset}\label{loading-and-understanding-the-dataset}

@@

\section{First look at the empirical
distributions}\label{first-look-at-the-empirical-distributions}

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-4-1} \end{center}

@@

\section{Looking at the islam-related
words}\label{looking-at-the-islam-related-words}

@@

@@

\begin{center}\includegraphics[width=1\linewidth]{_main_files/figure-latex/unnamed-chunk-8-1} \end{center}

@@

\section{Bayesian model structure and
assumptions}\label{bayesian-model-structure-and-assumptions}

@@

\section{Choosing predictors}\label{choosing-predictors}

@@

\section{Dataset-level coefficients}\label{dataset-level-coefficients}

@@

\section{Model structure and
assumptions}\label{model-structure-and-assumptions}

@@

\section{Protected classes in Reddit and Google
embeddings}\label{protected-classes-in-reddit-and-google-embeddings}

@@

\section{Dataset-level coefficients after
debiasing}\label{dataset-level-coefficients-after-debiasing}

@@

\section{Protected classes after
debiasing}\label{protected-classes-after-debiasing}

@@

\hypertarget{refs}{}
\hypertarget{ref-Bolukbasi2016Man}{}
Bolukbasi, T., Chang, K., Zou, J. Y., Saligrama, V., \& Kalai, A.
(2016). Man is to computer programmer as woman is to homemaker?
Debiasing word embeddings. \emph{CoRR}, \emph{abs/1607.06520}. Retrieved
from \url{http://arxiv.org/abs/1607.06520}

\hypertarget{ref-Caliskan2017Semantics}{}
Islam, A. C., Bryson, J. J., \& Narayanan, A. (2016). Semantics derived
automatically from language corpora necessarily contain human biases.
\emph{CoRR}, \emph{abs/1608.07187}. Retrieved from
\url{http://arxiv.org/abs/1608.07187}

\hypertarget{ref-manzini2019black}{}
Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
to criminal as caucasian is to police: Detecting and removing multiclass
bias in word embeddings.
